---
title: "Cell Type Annotation - T cell activation (rep 2)"
author: "Ramon Massoni-Badosa"
date: "12/4/2019"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction
The purpose of this notebook is to annotate the cell types of the two new T cell culturing & activation experiments (replicate 2). It will follow the same steps as replcate 1, so we refer to the notebook "03-annotation.Rmd" for a full explanation of each step.

# Pre-processing

## Package loading

```{r}
library(scater)
library(scran)
library(Seurat)
library(ggpubr)
library(kBET)
library(cluster)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(viridis)
library(tidyverse)
```

## Source script with functions

```{r}
source("bin/utils.R")
```

## Load data

```{r}
t_act_l <- readRDS("results/R_objects/t_act_Seurat_list_filtered_normalized.rds")
```

Let us filter the previous list to retain only replicate 2:

```{r}
t_act_l <- t_act_l[c("day_0_rep2", "day_1_rep2")]
```

# Find Variable Genes

```{r}
t_act_0_rep2 <- SplitObject(t_act_l$day_0_rep2, split.by = "donor")
t_act_1_rep2 <- SplitObject(t_act_l$day_1_rep2, split.by = "donor")
t_act_l <- list(
  t_act_0_rep2$Female2, 
  t_act_0_rep2$Female3, 
  t_act_1_rep2$Female2, 
  t_act_1_rep2$Female3
)
names(t_act_l) <- c("0_rep2_F2", "0_rep2_F3", "1_rep2_F2", "1_rep2_F3")
t_act_l <- purrr::map(t_act_l, FindVariableFeatures)
var_plots <- purrr::map(t_act_l, VariableFeaturePlot)
var_plots <- purrr::map2(t_act_l, var_plots, function(seurat, p) {
  LabelPoints(
  plot = p, 
  points =  head(VariableFeatures(seurat), 10), 
  repel = TRUE
  )
})
var_plots
```

Again, in the activated cells (day 1), we see that interferon gamma is the most variable gene.

# Scale data

```{r}
t_act_l <- purrr::map(t_act_l, ScaleData)
```

# Linear dimensionality reduction (PCA)

```{r}
t_act_l <- purrr::map(t_act_l, RunPCA)
purrr::map(t_act_l, VizDimLoadings, dims = 1:3, reduction = "pca")
purrr::map(t_act_l, DimPlot, reduction = "pca")
```
# Cluster cells

```{r}
t_act_l <- purrr::map(t_act_l, FindNeighbors, dims = 1:20)
resolutions <- c(
  0.005,
  seq(0.01, 0.1, by = 0.01),
  seq(0.1, 1, by = 0.1), seq(1, 10, by = 1)
)
avgs_sil_width_dfs <- purrr::map2(t_act_l, names(t_act_l), function(seurat, condition) {
  print(condition)
  avgs_sil_width_df <- data.frame(num_k = c(), sil_width = c(), resolution = c())
  num_k <- 1
  for (res in resolutions) {
    print(str_c("Current resolution is ", res))
    seurat <-  FindClusters(seurat, resolution = res, verbose = FALSE)
    curr_num_k <- length(levels(seurat$seurat_clusters))
    if (curr_num_k == num_k) {
      next
    } else {
      sil_width <- calc_sil_width(
        object = seurat, 
        clusters = seurat$seurat_clusters, 
        npcs = 5, 
        ncell = 2500
      )
      print(str_c("Current silhouette width is ", sil_width))
      curr_df <- data.frame(num_k = curr_num_k, sil_width = sil_width, resolution = res)
      avgs_sil_width_df <- rbind(avgs_sil_width_df, curr_df)
      num_k <- curr_num_k 
      print(str_c("Current number of clusters is ", num_k))
    }
  }
  avgs_sil_width_df
})
resolution_plots <- purrr::map2(avgs_sil_width_dfs, names(avgs_sil_width_dfs), function(df, cond) {
  ggplot(df, aes(num_k, sil_width, label = num_k)) +
    geom_text() +
    labs(title = cond,
         x = "Number of clusters (k)", 
         y = "Average Silhouette Width") +
    theme_classic() +
    theme(plot.title = element_text(hjust = 0.5))
})
resolution_plots
```

We aim to determine a number of clusters that maximizes silhouette width while preventing overstratification. Judgding by the previous plots, we define the following k:

* 0_rep2_F2: 5
* 0_rep2_F3: 4
* 1_rep2_F2: 5
* 1_rep2_F3: 5

```{r}
optimal_k <- c(5, 4, 5, 5)
t_act_l <- purrr::pmap(list(t_act_l, avgs_sil_width_dfs, optimal_k), function(seurat, df, opt_k) {
  opt_resolution <- df[df$num_k == opt_k, "resolution"]
  seurat <- FindClusters(object = seurat, resolution = opt_resolution)
  seurat
})
purrr::map(t_act_l, ~ length(levels(.x$seurat_clusters)))
```

# Non-linear dimensionality reduction 

```{r}
t_act_l <- purrr::map(t_act_l, function(seurat) {
  seurat@reductions$PCA <- NULL
  seurat@reductions$TSNE <- NULL
  seurat@reductions$UMAP <- NULL
  seurat
})
t_act_l <- purrr::map(t_act_l, RunTSNE, reduction = "pca", dims = 1:20)
t_act_l <- purrr::map(t_act_l, RunUMAP, reduction = "pca", dims = 1:20)
purrr::map(t_act_l, DimPlot, reduction = "tsne")
purrr::map(t_act_l, DimPlot, reduction = "umap")
```

# Find cluster markers

```{r}
markers_l <- purrr::map(t_act_l, FindAllMarkers)
saveRDS(markers_l, "results/R_objects/t_act_markers_list_rep2.rds")
marker_selection <- purrr::map2(t_act_l, markers_l, function(seurat, df) {
  num_k <- length(unique(seurat$seurat_clusters))
  selection <- unlist(purrr::map(0:(num_k-1), ~df[df$cluster == .x, "gene"][0:8]))
  selection
})
heatmaps_markers <- purrr::map2(t_act_l, marker_selection, ~DoHeatmap(.x, features = .y) + NoLegend())
heatmaps_markers
```

# Cell cycle scoring

```{r}
t_act_l <- purrr::map(t_act_l, function(seurat) {
  s_genes <- cc.genes$s.genes[cc.genes$s.genes %in% rownames(seurat@assays$RNA@scale.data)]
  g2m_genes <- cc.genes$g2m.genes[cc.genes$g2m.genes %in% rownames(seurat@assays$RNA@scale.data)]
  seurat <- CellCycleScoring(object = seurat, s.features = s_genes, g2m.features = g2m_genes)
  seurat
})
cycling_gg <- purrr::map(t_act_l, FeaturePlot, features = c("S.Score", "G2M.Score"), cols = viridis(20))
cycling_gg
```

# T cell stimulation signature

```{r}
anti_cd3_signature <- c("IL2RA", "IL1R", "IL12RB2", "IL18R1", "IL21R", "IL23R", 
                        "FOXP3", "CTLA4", "TNFRSF18", "LAG3", "PDCD1", "INFG", 
                        "IL17A", "IL17F", "LIF", "TNF", "IL1B", "IL6", "IL32", 
                        "IL2", "IL3", "IL9", "IL12B", "IL21", "IL22", "CD25", 
                        "IL1R2", "TBX21", "STAT4", "GMZB", "EOMES", "GZMB", 
                        "TNFRSF4", "TNFRSF9", "GITR", "XCL1", "IDO1", "NR4A1", 
                        "RORC")
t_act_l <- purrr::map(
  t_act_l, 
  AddModuleScore, 
  features = list(anti_cd3_signature), 
  name = "anti_cd3_score"
)
anti_cd3_gg <- purrr::map(
  t_act_l, 
  FeaturePlot, 
  features = "anti_cd3_score1", 
  reduction = "umap",
  cols = viridis(30),
  pt.size = 1.25
)
anti_cd3_gg
```

#  Annotation

## 0_rep2_F2
Cluster ID | Cell type       
-----------|---------------
0          | T 
1          | NK          
2          | Monocytes
3          | B 
4          | Dendritic Cells

```{r}
FeaturePlot(t_act_l$`0_rep2_F2`, features = c("IL7R", "CD3D"))
t_act_l$`0_rep2_F2`$cell_type <- case_when(
  t_act_l$`0_rep2_F2`$seurat_clusters == "0" ~ "T",
  t_act_l$`0_rep2_F2`$seurat_clusters == "1" ~ "NK",
  t_act_l$`0_rep2_F2`$seurat_clusters == "2" ~ "Monocyte",
  t_act_l$`0_rep2_F2`$seurat_clusters == "3" ~ "B",
  t_act_l$`0_rep2_F2`$seurat_clusters == "4" ~ "Dendritic Cell"
)
Idents(t_act_l$`0_rep2_F2`) <- "cell_type"
DimPlot(t_act_l$`0_rep2_F2`, reduction = "umap", label = TRUE) + NoLegend()
```

## 0_rep2_F3
Cluster ID | Cell type       
-----------|---------------
0          | CD4 T
1          | Monocytes 
2          | Cytotoxic
3          | B 

```{r}
FeaturePlot(t_act_l$`0_rep2_F3`, features = c("IL7R", "CD3D"))
t_act_l$`0_rep2_F3`$cell_type <- case_when(
  t_act_l$`0_rep2_F3`$seurat_clusters == "0" ~ "CD4 T",
  t_act_l$`0_rep2_F3`$seurat_clusters == "1" ~ "Monocyte",
  t_act_l$`0_rep2_F3`$seurat_clusters == "2" ~ "Cytotoxic",
  t_act_l$`0_rep2_F3`$seurat_clusters == "3" ~ "B"
)
Idents(t_act_l$`0_rep2_F3`) <- "cell_type"
DimPlot(t_act_l$`0_rep2_F3`, reduction = "umap", label = TRUE) + NoLegend()
```

## 1_rep2_F2
Cluster ID | Cell type       
-----------|---------------
0          | Cycling T
1          | NK          
2          | Activated T
3          | B 
4          | Mitotic

```{r}
FeaturePlot(t_act_l$`1_rep2_F2`, features = c("IL7R", "CD3D"))
t_act_l$`1_rep2_F2`$cell_type <- case_when(
  t_act_l$`1_rep2_F2`$seurat_clusters == "0" ~ "Cycling T",
  t_act_l$`1_rep2_F2`$seurat_clusters == "1" ~ "NK",
  t_act_l$`1_rep2_F2`$seurat_clusters == "2" ~ "Activated T",
  t_act_l$`1_rep2_F2`$seurat_clusters == "3" ~ "B",
  t_act_l$`1_rep2_F2`$seurat_clusters == "4" ~ "Mitotic"
)
Idents(t_act_l$`1_rep2_F2`) <- "cell_type"
DimPlot(t_act_l$`1_rep2_F2`, reduction = "umap", label = TRUE) + NoLegend()
```

## 1_rep2_F3
Cluster ID | Cell type       
-----------|---------------
0          | Cycling T
1          | Activated T          
2          | NK
3          | B 
4          | Mitotic

```{r}
FeaturePlot(t_act_l$`1_rep2_F3`, features = c("IL7R", "CD3D"))
t_act_l$`1_rep2_F3`$cell_type <- case_when(
  t_act_l$`1_rep2_F3`$seurat_clusters == "0" ~ "Cycling T",
  t_act_l$`1_rep2_F3`$seurat_clusters == "1" ~ "Activated T",
  t_act_l$`1_rep2_F3`$seurat_clusters == "2" ~ "NK",
  t_act_l$`1_rep2_F3`$seurat_clusters == "3" ~ "B",
  t_act_l$`1_rep2_F3`$seurat_clusters == "4" ~ "Mitotic"
)
Idents(t_act_l$`1_rep2_F3`) <- "cell_type"
DimPlot(t_act_l$`1_rep2_F3`, reduction = "umap", label = TRUE) + NoLegend()
```

# Save

```{r}
saveRDS(t_act_l, file = "results/R_objects/t_act_Seurat_list_annotated_rep2")
```

# Session Info

```{r}
sessionInfo()
```


