---
title: "Cell Culture and Activation - Replicate 2"
author: "Ramon Massoni-Badosa"
date: "12/4/2019"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction
The objective of this notebook is to reproduce the findings observed for the male donor in the previous notebook (04-cell_culture_activation_effect.Rmd). We found that the female donor in the previous analysis was invalid, so we had only one reliable donor. To finally conclude that the cold-shock transcriptomic signature is removed after culturing and activating PBMC with anti-CD3 antibodies, herein we will analyze data from two more donors (female2, female3). For reproducibility and consistency purposes, the analysis will be almost the same as in the previous notebook, so we refer to that one for a full explanation of each step. 

# Pre-processing

## Package loading

```{r}
library(scater)
library(scran)
library(Seurat)
library(ggpubr)
library(kBET)
library(cluster)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(viridis)
library(tidyverse)
```

## Source script with functions

```{r}
source("bin/utils.R")
```

## Load data

```{r}
t_act_l <- readRDS("results/R_objects/t_act_Seurat_list_annotated_rep2")
dea_list <- readRDS("../2-CLL/results/R_objects/dea_results_per_patient.rds")
```

# Effect Cell Culture

```{r fig.wide=TRUE}
t_act_l <- purrr::map(t_act_l, RunTSNE, reduction = "pca", dims = 1:15)
t_act_l <- purrr::map(t_act_l, function(seurat) {
  seurat$time <- factor(seurat$time, levels = c("0h", "8h", "24h"))
  Idents(seurat) <- "time"
  seurat
})
t_act_umaps <- purrr::map(t_act_l, function(seurat) {
  DimPlot(seurat, reduction = "umap", cols = c("#999999", "#58315E", "#D36A55"))
})
# saveRDS(t_act_umaps, "results/R_objects/ggplots/umaps_t_activation_rep2.rds")
t_act_tsnes <- purrr::map(t_act_l, function(seurat) {
  DimPlot(seurat, reduction = "tsne", cols = c("#999999", "#58315E", "#D36A55"))
})
# saveRDS(t_act_tsnes, "results/R_objects/ggplots/tsnes_t_activation_rep2.rds")

# View results
iterable <- list(c("0_rep2_F2", "1_rep2_F2"), c("0_rep2_F3", "1_rep2_F3"))
quali_intermix <- purrr::map(iterable, function(x) {
  plot_list <- t_act_tsnes[x]
  plot_list[[1]] <- plot_list[[1]] +
    ggtitle("Original") +
    theme(plot.title = element_text(hjust = 0.5, size = 13))
  plot_list[[2]] <- plot_list[[2]] +
    ggtitle("Cultured") +
    theme(plot.title = element_text(hjust = 0.5, size = 13))
  p <- ggarrange(
    plotlist = plot_list, 
    nrow = 1, 
    ncol = 2, 
    common.legend = TRUE, 
    legend = "right"
  )
  p
})
names(quali_intermix) <- c("female2", "female3")
quali_intermix
```

# Cold-shock Score

```{r}
# Find cold shock signature
cold_shock_signature <- readRDS("../1-PBMC/results/R_objects/cold_shock_signature_up.rds")

# Compute cold shock score
t_act_l <- purrr::map(t_act_l, function(seurat) {
  seurat <- AddModuleScore(
    object = seurat, 
    features = list(cold_shock_signature), 
    name = "cold_shock_score"
  )
  seurat
})

# Visualize cold-shock score
cold_shock_dfs <- purrr::map(iterable, function(x) {
  t_act_l[x] %>% 
    map(~ .x@meta.data) %>%
    map(~ .x %>% dplyr::select("day", "time", "cold_shock_score1")) %>% 
    bind_rows(.id = "condition") %>% 
    mutate(condition = ifelse(day == "day0", "Original", "Cultured")) %>% 
    mutate(condition = factor(condition, levels = c("Original", "Cultured")))
})
cold_shock_violins <- purrr::map(cold_shock_dfs, function(df) {
  df %>% 
    ggplot(aes(time, cold_shock_score1, fill = condition)) +
    geom_boxplot(outlier.size = 0.1) +
    labs(x = "", y = "Cold-Shock Score", fill = "") +
    scale_fill_manual(values = c("#3374A1", "#E1812C")) +
    theme_classic()
})
names(cold_shock_violins) <- c("female2", "female3")
cold_shock_violins
# saveRDS(cold_shock_violins, "results/R_objects/ggplots/boxplot_cold_shock_t_act_rep2.rds")
```

# Session Info

```{r}
sessionInfo()
```

