---
title: "Effect of Sampling Time"
author: "Ramon Massoni-Badosa"
date: "10/20/2019"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction
The objective of this notebook is to test if the time required to cryopreserve a blood sample is a source of technical artifacts in single-cell RNA-seq (scRNA-seq) data. We will do so by computing and visualizing the 'mixability' of PBMC that were kept at varying times (0h, 2h, 8h, 24h, 48h) prior to freezing.

## Package loading

```{r}
library(scater)
library(Seurat)
library(ggpubr)
library(purrr)
library(kBET)
library(tidyverse)
```            

## Source script with function definitions

```{r}
source("bin/utils.R")
```

## Load data
Our data consists of a list of two (male/female) demultiplexed, filtered, normalized and annotated Seurat objects. Also, we have a merged dataset that combines both male and female:

```{r}
pbmc_l <- readRDS("results/R_objects/10X_pbmc_Seurat_donors_list_clustered.RDS")
pbmc_l
pbmc <- readRDS("results/R_objects/10X_pbmc_Seurat_clustered.RDS")
pbmc <- subset(pbmc, subset = condition %in% c("0h", "2h", "8h", "24h_RT", "48h_RT"))
```

# Visualize mixability
Let us visualize the mixability of the PBMCs with different sampling times. Our hypothesis is that, if this variable introduces artifacts, cells will cluster by time. 

```{r}
# Recode variables
conds <- c("0h", "2h", "8h", "24h_RT", "48h_RT", "24h_4C", "48h_4C")
pbmc_l <- purrr::map(pbmc_l, function(s) {
  s$condition <- factor(s$condition, levels = conds)
  levels(s$condition) <- conds %>%
    str_remove("RT") %>%
    str_remove("_")
  s$temperature <- case_when(
    s$condition == "0h" ~ "fresh",
    s$condition %in% c("2h", "8h", "24h", "48h") ~ "RT",
    s$condition %in% c("24h4C", "48h4C") ~ "4ºC"
  )
  s <- subset(s, subset = temperature != "4ºC")
  s <- pre_process_seurat(s)
  s$time <- s$condition
  Idents(s) <- "time"
  s
})

# Visualize mixability (UMAP)
palette <- c("#999999", "#92e8df", "#632c63", "#e4624e", "#c0e212")
umaps_time <- purrr::map(
  pbmc_l, 
  DimPlot, 
  reduction = "umap", 
  cols = palette
)
umaps_time

# Visualize mixability (TSNE)
tsnes_time <- purrr::map(
  pbmc_l, 
  DimPlot, 
  reduction = "tsne", 
  cols = palette
)
tsnes_time
saveRDS(tsnes_time, "results/R_objects/ggplots/tsne_time_points_PBMC_Seuratv3.rds")
saveRDS(pbmc_l$male, "results/R_objects/pbmc_Seurat_male.rds")
```

As we can see, within each cell type cells cluster by time.

# Principal Component Analysis (PCA)
Let us asses if, for each cell type, time is the major source of variance in the dataset:

```{r}
pbmc_types <- SplitObject(pbmc, split.by = "cell_type")
pbmc_types <- purrr::map(pbmc_types, pre_process_seurat)
pbmc_types_pca <- purrr::map(pbmc_types, function(seurat) {
  df <- as.data.frame(seurat@reductions$pca@cell.embeddings[, 1:2])
  df$time <- seurat$condition
  df$time <- str_remove(df$time, "_RT")
  df$time <- factor(df$time, levels = c("0h", "2h", "8h", "24h", "48h"))
  df
})
pbmc_types_pca <- bind_rows(pbmc_types_pca, .id = "cell_type")
pbmc_types_pca$cell_type <- factor(
  pbmc_types_pca$cell_type, 
  levels = c("T", "NK", "Monocyte", "B")
)
palette2 <- c("#c20a35", "#aa2edc", "#71bdd0", "#bbaa2a")
pc1_vs_time <- pbmc_types_pca %>% 
  filter(PC_1 < 15) %>% 
  ggplot(aes(time, PC_1, fill = cell_type)) +
    geom_boxplot(outlier.shape = NA) +
    facet_grid(.~cell_type) +
    labs(x = "", y = "PC1") +
    scale_fill_manual("", values = palette2) +
    theme_classic() +
    theme(legend.position = "none",
          axis.title.y = element_text(size = 13, face = "bold"))
pc1_vs_time
saveRDS(pc1_vs_time, "results/R_objects/ggplots/pc1_time_Seuratv3_gg.rds")
```

# Zoom-in into specific cell types

```{r}
pbmc_types <- map(pbmc_types, function(seurat) {
  seurat$time <- seurat$condition
  seurat$time <- str_remove(seurat$time, "_RT")
  seurat$time <- factor(seurat$time, c("0h", "2h", "8h", "24h", "48h"))
  Idents(seurat) <- "time"
  seurat
})
pbmc_types <- map(pbmc_types, RunTSNE)
tsne_types <- map(pbmc_types, DimPlot, reduction = "tsne", cols = c("#999999", "#92e8df", "#632c63", "#e4624e", "#c0e212"))
saveRDS(tsne_types, "results/R_objects/ggplots/tsne_time_points_PBMC_cell_types_Seuratv3.rds")
```

# kBET

To quantify the mixability between time-points, we will use the metric [kBET](https://www.nature.com/articles/s41592-018-0254-1). Intuitively, kBET quantifies as well-mixed those datasets for which most subsets of neareast neighbors have the same local batch distribution as the global of the dataset. The documentation of the package can be found [here](https://github.com/theislab/kBET).

```{r}
# Calculate rejection rates
times <- c("2h", "8h", "24h", "48h")
rejection_rate_df <- purrr::map(pbmc_l, function(seurat) {
  Idents(seurat) <- "time"
  rejection_rate_dbl <- purrr::map_dbl(times, function(t) {
    seurat_sub <- subset(seurat, idents = c("0h", t))
    seurat_sub$time <- seurat_sub$condition
    seurat_sub$time <- str_remove(seurat_sub$time, "_RT")
    seurat_sub$is_biased <- ifelse(seurat_sub$time == "0h", FALSE, TRUE)
    Idents(seurat_sub) <- "time"
    mat <- as.matrix(seurat_sub@reductions$umap@cell.embeddings)
    batch <- seurat_sub$is_biased
    batch_estimate <- kBET(mat, batch = batch, verbose = TRUE, heuristic = FALSE, testSize = 50)
    batch_results <- batch_estimate$results
    batch_results <- batch_results %>% 
      mutate(cell_type = seurat_sub$cell_type,
             time = seurat_sub$time) %>% 
      filter(tested == 1) %>% 
      summarise(rejection_rate = 100 - mean(kBET.pvalue.test > 0.05) * 100)
    batch_results$rejection_rate
  })
  data.frame(time = times, rejection_rate = rejection_rate_dbl)
})

# Plot rejection rates
rejection_rate_gg <- purrr::map(rejection_rate_df, function(df) {
  df %>% 
    mutate(time = factor(time, levels = c("2h", "8h", "24h", "48h"))) %>% 
    ggplot(aes(time, rejection_rate)) + 
      geom_col() +
      labs(x = "", y = "kBET (rejection rate)") +
      theme_classic()
})
rejection_rate_gg
saveRDS(rejection_rate_gg, "results/R_objects/ggplots/barplot_rejection_rate_pbmc.rds")
```

# Session Information

```{r}
sessionInfo()
```

