---
title: "PBMC: Computational Correction Cold Shock"
author: "Ramon Massoni-Badosa"
date: "10/21/2019"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", fig.align='center', 
                      message=FALSE, warning = FALSE)
options(width = 1200)
```

# Introduction

The objective of this notebook is to create a cell-specific "cold-shock score" using the cold-shock transcriptomic signature. We will assess the potential of this score in predicting and correcting cold-shocked cells.

# Pre-processing

## Package loading

```{r}
library(scater)
library(scran)
library(Seurat)
library(ggpubr)
library(biomaRt)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(GOplot)
library(GOstats)
library(kBET)
library(topGO)
library(ggrepel)
library(viridis)
library(DescTools)
library(pheatmap)
library(tidyverse)
```

## Source script with functions

```{r}
source("bin/utils.R")
```

## Load data
We dispose of a demultiplexed, filtered, normalized and annotated Seurat object that contains the transcriptome and metadata  of >10,000 Peripheral Blood Mononuclear Cells (PBMC) from two donors (male/female) that were kept at varying times at RT (0h, 2h, 8h, 24h, 48h).

```{r}
pbmc <- readRDS("results/R_objects/10X_pbmc_Seurat_clustered.RDS")
pbmc <- subset(pbmc, subset = condition %in% c("0h", "2h", "8h", "24h_RT", "48h_RT"))
pbmc$time <- factor(pbmc$condition, levels = c("0h", "2h", "8h", "24h_RT", "48h_RT"))
levels(pbmc$time) <-  c("0h", "2h", "8h", "24h", "48h")
Idents(pbmc) <- "cell_type"
DimPlot(pbmc)
```

# Cold-shock score
We will divide the PBMC dataset in training set (male) and test set (female). First, let us find the signature in the training set:

```{r}
pbmc$is_biased <- ifelse(pbmc$time %in% c("0h", "2h"), "unbiased", "biased")
pbmc_donor <- SplitObject(pbmc, split.by = "donor")
Idents(pbmc_donor$male) <- "is_biased"
dea <- FindMarkers(pbmc_donor$male, ident.1 = "biased")
cold_shock_signature <- rownames(dea)[1:150]
head(cold_shock_signature)
```

With the signature, we can compute a cold-shock score for every cell in the test set:

```{r}
pbmc_donor <- map(pbmc_donor, pre_process_seurat)
pbmc_donor <- map(pbmc_donor, RunTSNE, reduction = "pca", dims = 1:15)
Idents(pbmc_donor$female) <- "time"
palette <- c("#999999", "#92e8df", "#632c63", "#e4624e", "#c0e212")
original_umap <- DimPlot(
  pbmc_donor$female, 
  reduction = "umap", 
  cols = palette
)
original_umap
original_tsne <- DimPlot(
  pbmc_donor$female, 
  reduction = "tsne", 
  cols = palette
)
pbmc_donor <- map(
  pbmc_donor, 
  AddModuleScore, 
  features = list(cold_shock_signature), 
  name = "cold_shock_score"
)
pbmc_donor <- purrr::map(pbmc_donor, function(seurat) {
  seurat$cold_shock_score1 <- -1 * seurat$cold_shock_score1
  seurat
})

vln_cold_score <- VlnPlot(
  pbmc_donor$female, 
  features = "cold_shock_score1", 
  cols = c("#999999", "#92e8df", "#632c63", "#e4624e", "#c0e212"),
  group.by = "time",
  pt.size = 0
)
vln_cold_score <- vln_cold_score +
  labs(title = "", x = "", y = "Cold-Shock Score") +
  theme(axis.text.x = element_text(angle = 0)) +
  theme(legend.position = "none")
vln_cold_score
# saveRDS(vln_cold_score, "results/R_objects/ggplots/violin_cold_shock_score.rds")
```

# Predict biased cells

To predict cold-shocked cells, we will use a logistic regression, using cold-shock score as explanatory variable and the probability of being cold-shocked as the response variable. To learn more about logistic regression and how to apply it in R, I recomend the book ["Introduction to Statistical Learning"](https://faculty.marshall.usc.edu/gareth-james/ISL/ISLR%20Seventh%20Printing.pdf), by Gareth James and colleagues.

```{r}
# Fit logistic regression model
set.seed(123)
random_genes <- sample(rownames(pbmc), size = length(cold_shock_signature), replace = FALSE)
pbmc_donor <- map(pbmc_donor, AddModuleScore, features = list(random_genes), name = "random_score")
train_df <- data.frame(
  cold_shock_score = pbmc_donor$male$cold_shock_score1,
  random_score = pbmc_donor$male$random_score1,
  class = factor(ifelse(pbmc_donor$male$is_biased == "biased", 1, 0))
)
logistic_fit <- glm(class ~ cold_shock_score, train_df, family = "binomial")
summary(logistic_fit)
logistic_random <- glm(class ~ random_score, train_df, family = "binomial")
summary(logistic_random)

# Predict probabilities on test set
test_df <- data.frame(
  cold_shock_score = pbmc_donor$female$cold_shock_score1,
  random_score = pbmc_donor$female$random_score1,
  class = factor(ifelse(pbmc_donor$female$is_biased == "biased", 1, 0))
)

test_df$probabilities_cold <- predict(logistic_fit, newdata = test_df, type = "response")
test_df$probabilities_random <- predict(logistic_random, newdata = test_df, type = "response")
plot(probabilities_cold ~ cold_shock_score, data = test_df)

# Plot ROC curve
thresholds <- seq(from = 0, to = 1, by = 0.01)
positives_ind <- which(test_df$class == 1)
negatives_ind <- which(test_df$class == 0)
accuracies_l <- map(c("probabilities_cold", "probabilities_random"), function(x) {
  l <- map(thresholds, function(thresh) {
    pred <- ifelse(test_df[[x]] > thresh, 1, 0)
    tpr <- mean(pred[positives_ind] == 1)
    fpr <- 1 - mean(pred[negatives_ind] == 0)
    list(tpr = tpr, fpr = fpr)
  })
  l
})
names(accuracies_l) <- c("cold_shock_score", "random_score")
accuracies_l <- map(accuracies_l, function(l) {
  tpr_dbl <- map_dbl(l, "tpr")
  fpr_dbl <- map_dbl(l, "fpr")
  roc_df <- data.frame(TPR = tpr_dbl, FPR = fpr_dbl)
  roc_df
})
accuracies_df <- bind_rows(accuracies_l, .id = "score")
roc_gg <- ggplot(accuracies_df, aes(FPR, TPR, color = score)) +
  geom_line() +
  labs(x = "False Positive Rate (FPR)", y = "True Positive Rate (TPR)", color = "") +
  scale_color_manual(values = c("limegreen", "darkgray"), labels = c("cold-shock", "random")) +
  theme_classic()
roc_gg
auroc <- map_dbl(c("cold_shock_score", "random_score"), function(x) {
  df <- accuracies_df[accuracies_df$score == x, ]
  AUC(x = df$FPR, y = df$TPR, method = "spline")
})
names(auroc) <- c("cold_shock_score", "random_score")
auroc
# saveRDS(roc_gg, file = "results/R_objects/ggplots/roc_curve_pbmc.rds")
```

# Regress Out
Now, let us regress the expression of each gene onto cold shock score. We will do this for each cell type separately to minimize [Simpson's paradox](https://ds.czbiohub.org/blog/Regression-Hazards/). Then, we will keep the residuals as the variability not explained by time:

```{r}
seu <- pbmc_donor$female
Idents(seu) <- "cell_type"
cell_types <- c("T", "Monocyte", "NK", "B")
mat_l <- purrr::map(cell_types, function(type) {
  print(type)
  seurat <- subset(seu, idents = type)
  mat <- as.matrix(seurat[["RNA"]]@data[VariableFeatures(seurat), ])
  lm_l <- apply(mat, 1, function(x) lm(x ~ seurat$cold_shock_score1))
  names(lm_l) <- VariableFeatures(seurat)
  residuals_list <- purrr::map(lm_l, "residuals")
  residuals_mat <- Reduce(rbind, residuals_list)
  rownames(residuals_mat) <- names(lm_l)
  colnames(residuals_mat) <- colnames(seurat)
  mean_expr <- rowMeans(mat)
  residuals_mat <- residuals_mat + mean_expr
  residuals_mat_sc <- scale(residuals_mat, center = TRUE, scale = TRUE)
  residuals_mat_sc <- residuals_mat_sc[rownames(seurat[["RNA"]]@scale.data), ]
})
names(mat_l) <- cell_types

# Join matrices and process
pbmc_female2 <- pbmc_donor$female
mat_all <- Reduce(cbind, mat_l)
mat_all <- mat_all[rownames(seu[["RNA"]]@scale.data), colnames(seu)]
pbmc_female2[["RNA"]]@scale.data <- mat_all
pbmc_female2 <- RunPCA(pbmc_female2)
pbmc_female2 <- RunUMAP(pbmc_female2, reduction = "pca", dims = 1:20)
pbmc_female2 <- RunTSNE(pbmc_female2, reduction = "pca", dims = 1:20)
Idents(pbmc_female2) <- "time"
umap_corrected <- DimPlot(pbmc_female2, reduction = "umap", cols = palette)
tsne_corrected <- DimPlot(pbmc_female2, reduction = "tsne", cols = palette)
umap_corrected
tsne_corrected
# ggarrange(plotlist = list(original_umap, umap_corrected), ncol = 2)
# saveRDS(
#   list(original = original_umap, corrected = umap_corrected), 
#   "results/R_objects/ggplots/umaps_original_correction_pbmc.rds"
# )
# saveRDS(
#   list(original = original_tsne, corrected = tsne_corrected), 
#   "results/R_objects/ggplots/tsnes_original_correction_pbmc.rds"
# )
```

# kBET
To quantify the mixability between time-points, we will use the metric [kBET](https://www.nature.com/articles/s41592-018-0254-1). Intuitively, kBET compares the proportion of k-nearest neighbors that belong to each time-point with the global distribution. We say that a dataset is well-mixed if the local neighborhood approximates the global. The documentation of the package can be found [here](https://github.com/theislab/kBET).

```{r}
# Calculate rejection rates
pbmc_female <- pbmc_donor$female
times <- c("2h", "8h", "24h", "48h")
acceptance_rates <- purrr::map(list(pbmc_female, pbmc_female2), function(seurat) {
  Idents(seurat) <- "time"
  acceptance_rate_dbl <- purrr::map_dbl(times, function(t) {
    seurat_sub <- subset(seurat, idents = c("0h", t))
    seurat_sub$is_biased <- ifelse(seurat_sub$time == "0h", FALSE, TRUE)
    mat <- as.matrix(seurat_sub@reductions$umap@cell.embeddings)
    batch <- seurat_sub$is_biased
    batch_estimate <- kBET(mat, batch = batch, verbose = TRUE, heuristic = FALSE, testSize = 50)
    batch_results <- batch_estimate$results
    batch_results <- batch_results %>% 
      mutate(cell_type = seurat_sub$cell_type,
             time = seurat_sub$time) %>% 
      filter(tested == 1) %>% 
      summarise(acceptance_rate = mean(kBET.pvalue.test > 0.05) * 100)
    batch_results$acceptance_rate
  })
  data.frame(time = times, acceptance_rate = acceptance_rate_dbl)
})
names(acceptance_rates) <- c("original", "corrected")
acceptance_rates <- bind_rows(acceptance_rates, .id = "is_corrected")
acceptance_rates_gg <- acceptance_rates %>% 
  mutate(time = factor(time, levels = c("0h", "2h", "8h", "24h", "48h")),
         is_corrected = factor(is_corrected, levels = c("original", "corrected"))) %>% 
  ggplot(aes(time, acceptance_rate, fill = is_corrected)) +
    geom_col(position = "dodge") +
    scale_y_continuous(limits = c(0, 100)) +
    scale_fill_manual(values = c("azure3", "chartreuse2")) +
    labs(x = "", y = "kBET (acceptance rate)", fill = "") +
    theme_classic() +
    theme(axis.text = element_text(size = 11), 
          axis.title.y = element_text(size = 13), 
          legend.text = element_text(size = 11))
acceptance_rates_gg
# saveRDS(acceptance_rates_gg, "results/R_objects/ggplots/barplot_acceptance_rate_correction_pbmc.rds")
```


# Session Information

```{r}
sessionInfo()
```

